---
title: "Data Project"
author: "Mingi Kang"
date: "2024-05-14"
output: pdf_document
---


# Provisional Deaths in California Counties by Demographic and Year 
### Years: 2019 - 2022

## Introduction 

My interest in this data project comes from my home city of Los Angeles. Los Angeles is the largest city in California by population and the second largest city in the United States. As of 2023, the population of California is approximately 39.1 million people. With this many people in mind, there are different cases for deaths in the state. Each county has a different demographic and being in one demographic county can have more significance to death by a certain reason.Among the residents of California, this data project aims to find the fraction of deaths that come from a particular type of cause based on the different demographic factors. 


## Data Collection | Data Cleaning 

** The demographic data is from 2023, and the total number of deaths by different causes of deaths in the counties are from 2021, we could not find the number of deaths in 2023, but since the demographic data is accurate up to 5 years, the results should not be significantly different if we were to use 2023 total number of deaths by different causes of deaths data. 




Before the analysis of the data, there had to be a couple of changes to the data file for final deaths by month by occurrence by county. Since there are 12 months and 3 years in this data set, I believe that it would have been better if all of the months combined into one year. Each county have 10s of deaths and by combining them, we can analyze the final deaths by year by occurrence by county instead of by month. 

# ```{r setup, include=FALSE}
# knitr::opts_chunk$set(warning = FALSE, message = FALSE, results='hide', echo = FALSE)
# ```

### Useful function for later 
```{R}
# Varius function for analysis 

# function to calculate log likelihood for a linear model 
calc.llk.lm <- function(lm.in){
  eps <- lm.in$residuals
  N <- length(eps) 
  var <- 1/(N) *sum(eps^2, na.rm=TRUE)
  llk <- sum(dnorm(eps, 0, sqrt(var), log=TRUE), na.rm=TRUE)
  return (llk) 
}

# Function to calculate the AIC
calc.aic <- function(model){
  # Getting number of parameters 
  K <-length(coef(model)) + 1 
  
  
  # Getting the log likelihood
  llk <- calc.llk.lm(model)
  
  aic <- 2 * K - 2 * llk 
  return (aic)
}

## Optional : Function to calculate the number of parameters in a linear model 
calc.num.parameters <- function(lm.in){
  K <-length(coef(lm.in)) + 1 
  return (K)
}


# Function to generate all possible subsets of a vector
getSubsets <- function(vec) {
  n <- length(vec)
  subsets <- list()
  for (i in 0:n) {
    subsets <- c(subsets, combn(vec, i, simplify = FALSE))
  }
  return(subsets)
}

# Function to generate all possible linear models
generateModels <- function(response, predictors, data) {
  # Get all subsets of predictors
  attach(data)
  predictorSubsets <- getSubsets(predictors)
  
  # Initialize a list to store the models
  models <- list()
  
  # For each subset of predictors...
  for (i in 1:length(predictorSubsets)) {
    # Skip if the subset is empty
    if (length(predictorSubsets[[i]]) == 0) next
    
    # Generate the formula for the linear model
    formula <- as.formula(paste(response, "~", paste(predictorSubsets[[i]], collapse = "+")))
    
    # Fit the linear model and store it in the list
    model <- try(lm(formula, data=data), silent=TRUE)
    if (!inherits(model, "try-error")) {
      models[[length(models) + 1]] <- model
    }
  }
  
  return(models)
}

```


### Loading up the data and attaching variables 
```{R}
# Load the data for demographic and death data 
data.set <- read.csv("Data/Final Data/demographic_death.csv")

# attaching variable names to columns
column.names <- c("County", "Population", 
                    "American-Indian", "Asian", "Black", "Hispanic", "Multi-Racial", "Pacific-Islander", "White", 
                    "0-5", "6-17", "18-64", "65+", 
                    "All-Causes", "Alzheimer-Disease", "Malignant-Neoplasms", 
                    "Chronic-Lower-Respiratory-Disease", "Diabetes-Mellitus", "Assult-Homocide",
                    "Disease-of-Heart", 
                    "Essential-Hypertension-Hypertensive-Renal-Disease", "Accident-Unintentional", 
                    "Chronic-Liver-Disease-Cirrhosis", "Nephritis-Nephrotic-Syndrome-Nephrosis",
                    "Parkinsons-Disease", "Influenza-Pneumonia", 
                    "Cerebrovascular-Disease", "Intentional-Self-Harm-Suicide")

# Attaching variable names to columns
names(data.set) <- c("C", "P", "AI", "A", "B", "H", "MR", "HPI", "W", 
                     "X0_5", "X6_17", "X18_64", "X65_plus", 
                     "AC", "AD", "MN", "CLRD", "DM", "AH", "DOH", "EHHRD", "AU", "CLDC", 
                     "NNSN", "PD", "IP", "CD", "ISHS")

# Demographic variables
dem <- c("P", "AI", "A", "B", "H", "MR", "HPI", "W", 
         "X0_5", "X6_17", "X18_64", "X65_plus")

# Cause of death variables
dea <- c("AC", "AD", "MN", "CLRD", "DM", "AH", "DOH", "EHHRD", "AU", "CLDC", 
         "NNSN", "PD", "IP", "CD", "ISHS")


# Standarizing the data 
std.data.set <- data.set 
for (i in 2:ncol(data.set)) {
  std.data.set[, i] <- (data.set[, i] - mean(data.set[,i], na.rm=TRUE))
  std.data.set[, i] <- std.data.set[,i]/sd(data.set[,i], na.rm=TRUE)
}
attach(std.data.set)

#### data.set -> original data
#### std.data.set -> standardized data 
```

## Calculating LLK and AIC for linear models 


```{R}
# All Causes
library(ggplot2)
all.causes <- dea[1]
regressors <- dem 
models <- generateModels(all.causes, regressors, data.set)

llk <- sapply(models, calc.llk.lm)
aic <- sapply(models, calc.aic)

# Create a data frame
df <- data.frame(
  Regressor = seq_along(aic),    # Create an index for the x-axis
  LLK = llk,                         # The LLK values for the y-axis 
  AIC = aic                         # The AIC values for the y-axis
)



# Plot the line graph for LLK 
ggplot(df, aes(x = Regressor, y = LLK)) +
  geom_point(color = "red") +  # Adding points to the line for better visibility
  labs(title = "Line Graph of LLK with Regressors",
       x = "Regressor Index",
       y = "LLK Value") +
  theme_minimal()

# Plot the line graph for AIC 
ggplot(df, aes(x = Regressor, y = AIC)) +
  geom_point(color = "red") +  # Adding points to the line for better visibility
  labs(title = "Line Graph of AIC with Regressors",
       x = "Regressor Index",
       y = "AIC Value") +
  theme_minimal()

```





```{R}
# Get LLK and AIC for linear models 
lm.matrix <- matrix(nrow = length(dea), ncol = 5)
std.matrix <- matrix(nrow = length(dea), ncol = 5)

all.lm <- list()
```

```{R}
for (i in 1:length(dea)) {
  dep.var <- dea[i]
  regressors <- dem 

  # Generate all possible models for data + standardized data
  models <- generateModels(dep.var, regressors, data.set)
  std.models <- generateModels(dep.var, regressors, std.data.set)

  # save models for later use 
  all.lm[[i]] <- models

  # Calculate LLK for each model 
  llk <- sapply(models, calc.llk.lm)
  std.llk <- sapply(std.models, calc.llk.lm)

  # Calculate AIC for each model
  aic <- sapply(models, calc.aic)
  std.aic <- sapply(std.models, calc.aic)

  # Store the results in the matrix
  maxLLKIndex <- which.max(llk)
  minAICIndex <- which.min(aic)
  std.maxLLKIndex <- which.max(std.llk)
  std.minAICIndex <- which.min(std.aic)

  # store them in model 
  lm.matrix[i, 1] <- dep.var
  lm.matrix[i, 2] <- maxLLKIndex
  lm.matrix[i, 3] <- max(llk)
  lm.matrix[i, 4] <- minAICIndex
  lm.matrix[i, 5] <- min(aic)

  std.matrix[i, 1] <- dep.var
  std.matrix[i, 2] <- std.maxLLKIndex
  std.matrix[i, 3] <- max(std.llk)
  std.matrix[i, 4] <- std.minAICIndex
  std.matrix[i, 5] <- min(std.aic)
  print("-\n")
}


lm.df <- as.data.frame(lm.matrix)
colnames(lm.df) <- c("Cause Death", "Max LLK Index", "Max LLK", "Min AIC Index", "Min AIC")

std.lm.df <- as.data.frame(std.matrix)
colnames(std.lm.df) <- c("Cause Death", "Max LLK Index", "Max LLK", "Min AIC Index", "Min AIC")

print(lm.df)
print(std.lm.df)

```

#### best models 
```{R}

```

```{R}
# 1. All Causes 
min.aic.lm <- all.lm[[1]][[3523]]
print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 2. "Alzheimer-Disease"
min.aic.lm <- all.lm[[2]][[3799]]
print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 3. Malignant Neoplasms
min.aic.lm <- all.lm[[3]][[1807]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 4. Chronic lower respiratory disease  
min.aic.lm <- all.lm[[4]][[3928]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```


```{R}
# 5. Diabetes Mellitus 
min.aic.lm <- all.lm[[5]][[3864]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```


```{R}
# 6. Assult Homicide
min.aic.lm <- all.lm[[6]][[3917]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```


```{R}
# 7. Disease of hear  
min.aic.lm <- all.lm[[7]][[3921]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```


```{R}
# 8. Essential Hypertension 
min.aic.lm <- all.lm[[8]][[4063]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```


```{R}
# 9. Accident Unintentional 
min.aic.lm <- all.lm[[9]][[1404]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 10. Chronic Liver 
min.aic.lm <- all.lm[[10]][[3800]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 11. Nephritis
min.aic.lm <- all.lm[[11]][[1021]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 12. Parkinsons
min.aic.lm <- all.lm[[12]][[1461]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 13. Influenza
min.aic.lm <- all.lm[[13]][[1535]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 14. Cerebrovascular
min.aic.lm <- all.lm[[14]][[706]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

```{R}
# 15. Intentional 
min.aic.lm <- all.lm[[15]][[791]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
```

### Analysis and Bootstrap for Malignant Neoplasms
```{R}
# 3. Malignant Neoplasms
min.aic.lm <- all.lm[[3]][[1807]]
summary(min.aic.lm)
print(min.aic.lm$coefficients)


# 65 + model 
# Assuming your data frame is named data.set and the variables are properly named
lm.65.plus <- lm(data.set$MN ~ data.set$P + data.set$A + data.set$B + data.set$H + data.set$W + data.set$X65_plus)

# Display the summary of the linear model
summary(lm.65.plus)
print(lm.65.plus$coefficients)

# comparing aic
min.aic.lm.aic <- calc.aic(min.aic.lm)
lm.65.plus.aic <- calc.aic(lm.65.plus)
cat("The AIC with X0_5: ", min.aic.lm.aic, "\n")
cat("The AIC with X65_plus: ", lm.65.plus.aic, "\n")

min.aic.lm.r.sq <- summary(min.aic.lm)$r.squared
lm.65.plus.r.sq <- summary(lm.65.plus)$r.squared
cat("R Squared \n") 
cat("The R Squared with X0_5: ", min.aic.lm.r.sq, "\n")
cat("The R Squared with X65_plus: ", lm.65.plus.r.sq, "\n")

lm.65.plus$coefficients[7]
min.aic.lm$coefficients[7]
```


```{R}
# Bootstrapping 
M <- 3000
X.0.5.coef <- rep(NA, M)
X.65.plus.coef <- rep(NA, M)

for (j in 1:M){
  
  bs.pos <- sample(1:N, N, replace=TRUE)
  df.bs <- data.set[bs.pos, ]
  
  ## Recalculate 
  lm.0.5.temp <- lm(df.bs$MN ~ df.bs$P + df.bs$A + df.bs$B + df.bs$H + df.bs$W + df.bs$X0_5)
  lm.65.plus.temp <- lm(df.bs$MN ~ df.bs$P + df.bs$A + df.bs$B + df.bs$H + df.bs$W + df.bs$X65_plus)

  X.0.5.coef[j] <- lm.0.5.temp$coefficients[7]
  X.65.plus.coef[j] <- lm.65.plus.temp$coefficients[7]
  
}

c.i.99.0.5 <- quantile(X.0.5.coef, c(0.005, 0.995))

c.i.99.65.plus <-quantile(X.65.plus.coef, c(0.005, 0.995))

cat("The 99th confidence interval for coefficient for age 0-5 is: ", c.i.99.0.5, "\n")
cat("The 99th confidence interval for coefficient for age 65+ is: ", c.i.99.65.plus, "\n")

```

### Analysis and Bootstrap for Intentional Suicide 

```{R}
# 15. Intentional 
min.aic.lm <- all.lm[[15]][[791]]

print(summary(min.aic.lm))
print(names(min.aic.lm$coefficients))
print(coef(min.aic.lm))
print(min.aic.lm$coefficients[2])

# Bootstrapping 
M <- 3000
W.coeff <- rep(NA, M)

for (j in 1:M){
  
  bs.pos <- sample(1:N, N, replace=TRUE)
  df.bs <- data.set[bs.pos, ]
  
  ## Recalculate 
  lm.temp <- lm(df.bs$ISHS ~ df.bs$W + df.bs$X0_5 + df.bs$X18_64 + df.bs$X65_plus)

  # Store coefficients
  W.coeff[j] <- lm.temp$coefficients[2]
}

c.i.99.white <- quantile(W.coeff, c(0.005, 0.995))
cat("The 99th confidence interval for coefficient for white is: ", 
    c.i.99.white , "\n")


```

### Correlation Matrix 
```{R}
### Correlation matrix 
library(corrplot)
cor.mat <- cor(data.set[, -1])
corrplot(cor.mat)
```

```{R}

```


## Plots for Heat Map of California 
```{r}
# total population based on the counties
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(viridis)

options(tigris_class = "sf")

# Get ca counties information 
ca_counties <- counties(state = "CA", cb = TRUE, class = "sf")

# order the counties in alphabetical along 
ca_counties <- arrange(ca_counties, NAME)

# add population data to the ca_counties data frame
ca_counties <- ca_counties %>%
  mutate(population = data.set$P)

# Create the heat map
ggplot(data = ca_counties) +
  geom_sf(aes(fill = population), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", name = "Population") +
  labs(title = "Population Heat Map of California Counties") +
  theme_minimal()


```

```{r}
# total Death by Alzheimer Disease based on the counties
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(viridis)

options(tigris_class = "sf")

# Get ca counties information 
ca_counties <- counties(state = "CA", cb = TRUE, class = "sf")

# order the counties in alphabetical along 
ca_counties <- arrange(ca_counties, NAME)

# add population data to the ca_counties data frame
ca_counties <- ca_counties %>%
  mutate(population = data.set$AD)

# Create the heat map
ggplot(data = ca_counties) +
  geom_sf(aes(fill = population), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", name = "Population") +
  labs(title = "Death by Alzheimer's Disease Heat Map of California Counties") +
  theme_minimal()

```

```{r}
# total population of White Californians based on the counties
library(ggplot2)
library(sf)
library(tigris)
library(dplyr)
library(viridis)

options(tigris_class = "sf")

# Get ca counties information 
ca_counties <- counties(state = "CA", cb = TRUE, class = "sf")

# order the counties in alphabetical along 
ca_counties <- arrange(ca_counties, NAME)

# add population data to the ca_counties data frame
ca_counties <- ca_counties %>%
  mutate(population = data.set$W)

# Create the heat map
ggplot(data = ca_counties) +
  geom_sf(aes(fill = population), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "magma", name = "Population") +
  labs(title = "White Population Heat Map of California Counties") +
  theme_minimal()

```


## Plots for Linear model and Standarized Linear Model 

